AbxD01 Analysis

### 2015-01-22 abxD01_analysis.Rmd
fixed min a bun in fig 1 graph, added alternative

This figure was built using the following R code.  The input files are listed in Github.

```{r, eval=FALSE}
#################################################
#Purpose: Convert csv matrix of columns you want converted into barcharts
#input: csv file following this heading format format:  sampleID, expgroup, nextDayCFU, Otu001, Otu002, Otu003, etc
#ex:file<-read.csv("~/Desktop/mothur/abxD01/barcharts/abxD01.final.tx.5.subsample.relabund.topdose.forlogscale2.csv
#Note: It is assumed that anything after the expgroup column are vectors containing data to be split by the groups in 'expgroup' 
#      and averaged with SD measured. 
#Output: matrices avg[],  std[], bp[y] with the barplot coordinates info stored for each subgraph
#################################################
# Parameters to change:
# CSV file: Group  expgroup  Otu001... (limited by most abund, end with 'Other', OTUs normalized +0.0001, expgroups #'d by graph order & sorted by first graph)
file<-read.csv("/Documents/Github/abxD01/Figure 1/abxD01.final.tx.1.subsample.relabund.topdose2.forlogscale.csv", header=T)
fileIDS<-read.csv("/Documents/Github/abxD01/Figure 1/topdose_tx1_barchart_ids.csv", header=T)
# Y Labels for each graph: 
abx<-c("Untreated", "Ciprofloxacin", "Clindamycin", "Vancomycin", "Streptomycin", "Cefoperazone", "Ampicillin", "Metronidazole")

# If you want each OTU on the Y axis to be sorted by the most abundant phylum and then decreasing abundance within that phylum, change to TRUE
# The default is false, which means sort be decreasing relative abundance in the top group (untreated/control)
sortbyphyl<-TRUE

# Highlight all and run!
#################################################

file<-file[,-1] #delete the first col of group names
avgs=NULL
avgs <-data.frame(levels(file$expgroup))
colnames(avgs) <- c("expgroup")
stds=NULL
stds <-data.frame(levels(file$expgroup))
colnames(stds) <- c("expgroup")

#Calculate average cdiff
avgCD=NULL
sdCD=NULL
avgCD<-tapply(file$nextDayCFU, file$expgroup, mean)
avgCD<-format(avgCD, scientific=TRUE, digits=2)
sdCD<-tapply(file$nextDayCFU, file$expgroup, sd)
file<-file[,-2] #delete the nexDayCFU column

for(i in 1:ncol(file)){

  if(i==1){
    ids<-names(file)
  }
  else{
    columni <- data.frame(file[,i])
    colnames(columni) <- c("columndata")

    #calculates mean for each column in file by expgroup and stores it in avgs data frame
    poop<- aggregate(columni$columndata~file$expgroup, FUN=mean) #poop stores the aggregated means for each group for this specific column
    colnames(poop) <- c(ids[1], ids[i])

    avgs<- merge(avgs, poop, by ="expgroup")

    #calculates st dev for each column in file by expgroup and stores it in stds data frame
    poop2<- aggregate(columni$columndata~file$expgroup, FUN=sd) #poop stores the aggregated means for each group for this specific column
    colnames(poop2) <- c(ids[1], ids[i])

    stds<- merge(stds, poop2, by ="expgroup")
  }
}

ordIDS <- fileIDS[order(fileIDS[,2]), ] #sort by the phyla
totalOTU <- dim(ordIDS)[1]

########for doing sort by phylum then within phylum (by relabund)
if(sortbyphyl == TRUE){
  ordIDS<-cbind(ordIDS, 0)

  for (i in 1:totalOTU){ #loop to fill out the final column for the phyla groups, while phyla is sorted!!
    if(i==1){
      ordIDS[i, 4] <- 1
    }
    else{
      if(ordIDS[i, 2]==ordIDS[i-1, 2]){
        ordIDS[i, 4] <- ordIDS[i-1, 4]
      }
      else{
        ordIDS[i, 4] <- ordIDS[i-1, 4] + 1
      }

    }
  }

  names(ordIDS)[4]<-paste("phynum")

  ##add row for phylum number to the avgs?
  #avgs<-rbind(avgs, 0)[c(9, 1, 2, 3, 4, 5, 6, 7, 8),] #this is hard coded for the top doses... also dont rerun this multiple times!!! might need to change
  avgs<-rbind(0,avgs) #this is hard coded for the top doses... also dont rerun this multiple times!!! might need to change, will give error

  numphyla<-max(ordIDS[,4])

  #test_avg <- avgs[, rev(order(avgs[1,2:length(avgs)-1]))] #sort all but other column (last)

  physums <- data.frame(c(1:numphyla))
  physums <- cbind(physums, 0)
  colnames(physums) <- c("Num", "Sum")

  leng<-length(avgs)

  #jtime<-NULL
  #itime<-NULL
  #inside<-FALSE
  #count<-0
  j<-1
  ##This loop is to fill the physums data frame with the sum of the relabund for each phylum represented
  for(i in 2:(leng-1)){ #searching through avgs, so leng is number of otus, -1 because i dont want to deal with "Other" column
    found<-FALSE
    #itime<-rbind(itime, i)
    j<-1
    while(found==FALSE){ #searching through ordIDS
      test<-ordIDS[j, 1] == names(avgs)[i]
      if(test) { 
        pnum<-ordIDS[j, 4]
        found<-TRUE
        inside<-TRUE
        avgs[1,i]<-pnum
      }
      else{
        j <- j + 1
        #count<-count+1
        #jtime<-rbind(jtime, j)

      }
    } #WARNING: could cause an error if dont find the ID, theres no check for the end of the list

    physums[pnum, 2] <- physums[pnum, 2] + avgs[2, i] ##add the relabund of the otu   
  }

  y<-rev(order(physums$Sum))
  x<-avgs[1, 2:(leng-1)]

  ord_physums<-physums[rev(order(physums[,2])),]
  sort_avgs<-data.frame(avgs$expgroup)
  names(sort_avgs)[1]<-c("expgroup")

  for(i in 1:numphyla){ #loop is to sort by decreasing phyla and perform internal sort within phylum, return the new sorted avgs to go into mavgs
    val<-NULL
    phy<-which(x %in% c(ord_physums[i,1])) #phy returns the indices for where that phyNum is
    for(j in 1:length(phy)){ #using locations of phylum i to find in avgs
      val[j]<-avgs[2,phy[j]+1]
    }
    phy<-phy[rev(order(val))]

    for(j in 1:length(phy)){ #now loop through the sorted sub cols and add them to the sort_avgs
      sort_avgs<-cbind(sort_avgs, avgs[,phy[j]+1])
      names(sort_avgs)[length(sort_avgs)]<-names(avgs)[phy[j]+1]
    }
  }

  attach(avgs)
  sort_avgs<-cbind(sort_avgs, Other)
  detach(avgs)

  sort_avgs <- sort_avgs[-1,] #get rid of phylum first row from earlier sorting

  #loop for getting the right ids in the right order as sort
  ids<-data.frame(1:(totalOTU))

  row.names(sort_avgs)<-sort_avgs$expgroup
  sort_avgs<-sort_avgs[,-1]

  for (i in 1:(length(sort_avgs)-1)){
    found<-FALSE
    #itime<-rbind(itime, i)
    j<-1
    while(found==FALSE){ #searching through ordIDS
      test<-ordIDS[j, 1] == names(sort_avgs)[i]
      if(test) { 
        ids[i,2]<-ordIDS[j, 3]
        found<-TRUE
      }
      else{
        j <- j + 1
        #count<-count+1
        #jtime<-rbind(jtime, j)
      }
    }

  }

  names(ids)<-c("x", "name")
  ids<-rbind(ids, data.frame(x=totalOTU+1, name= "Other"))

  mavgs<-as.matrix(sort_avgs)

  leng<-dim(mavgs)[2]

  numgr <- nlevels(file$expgroup)
} ##end if(sortbyphyl == TRUE){

####order the results by the first group alpha numerically
if(sortbyphyl == FALSE){

  row.names(avgs)<-avgs$expgroup
  avgs<-avgs[,-1]
  ordered_avgs<- avgs[, rev(order(avgs[1,1:length(avgs)-1]))] #sort all but other column (last)
  attach(avgs)
  ordered_avgs <- cbind(ordered_avgs, Other) #put other back on
  detach(avgs)

  mavgs<-as.matrix(ordered_avgs)
  leng<-dim(mavgs)[2]

  numgr <- nlevels(file$expgroup)

  ids<-data.frame(1:(totalOTU))

  for (i in 1:(length(ordered_avgs)-1)){
    found<-FALSE
    #itime<-rbind(itime, i)
    j<-1
    while(found==FALSE){ #searching through ordIDS
      test<-ordIDS[j, 1] == names(ordered_avgs)[i]
      if(test) { 
        ids[i,2]<-ordIDS[j, 3]
        found<-TRUE
      }
      else{
        j <- j + 1
        #count<-count+1
        #jtime<-rbind(jtime, j)
      }
    }

  }

  names(ids)<-c("x", "name")
  ids<-rbind(ids, data.frame(x=totalOTU+1, name= "Other"))

}  #end if(sortbyphyl == FALSE){

###started to do the stds table... didnt complete
#stds<-stds[,-1]
#mstds<-as.matrix(stds)

###PLOT PARAMETERS
par(mfrow=c(numgr+1, 1)) #+1 to give extra labeling space
par(mar=c(0.3, 8, 0.5, 2) +0.1, mgp=c(4.5, 1, 0)) #default is 5.1 4.1 4.1 2.1, bot/left/top/right, also default mgp is c(3,1,0)

for(j in 1:numgr){

  if(j != numgr){
    barplot(mavgs[j, 1:leng], ylab=NULL, col="black", yaxt="n", xaxt="n", ylim=c(0.0001, 1), log="y", cex.names=3)
    #error.bar(bp[k], mavgs[j, 1:leng[2]], mstds[j, 1:leng[2]])  #the bp[k] was for storing the barplot locations to use for these errors
    # k <- k+1
    axis(2, las=1, at=c(.0001, .001, .01, .1, 1), labels=c(0, .001, .01, .1, 1), cex.axis=1.1)
    mtext(abx[j], side=2, line=6, cex=.8)
    mtext(avgCD[j], side=2, line=4.5, cex=.8)
    abline(h=c(0.0001, 1), lwd=3)
    abline(h=c(0.001, 0.01, 0.1), col="black", lty="longdash", lwd=1.5)
    abline(h=c(.25,.5, .75), col="black", lty="dashed")
  }

  else{ #the last graph needs different margins    
    label<-barplot(mavgs[j, 1:leng], col="black",ylab=NULL,  yaxt="n", xaxt="n", ylim=c(0.0001, 1), log="y", cex.names=3)
    #error.bar(bp[k], mavgs[j, 1:leng[2]], mstds[j, 1:leng[2]])  #the bp[k] was for storing the barplot locations to use for these errors
    axis(2, las=1, at=c(.0001, .001, .01, .1, 1), labels=c(0, .001, .01, .1, 1), cex.axis=1.1)
    mtext(abx[j], side=2, line=6, cex=.8)
    mtext(avgCD[j], side=2, line=4.5, cex=.8)
    axis(1, at=(label[,1]), labels=FALSE)
    text(label[,1]+.13, .00005, label=ids[,2], xpd=NA, pos=2, srt=45, cex=1.2)
    #text(-3.9,.0001, label=expression(paste(italic("C.d."), " CFU/g Feces:")), xpd=NA, pos=2, srt=90, cex=1.1)
    abline(h=c(0.0001,1), lwd=3)
    abline(h=c(0.001, 0.01, 0.1), col="black", lty="longdash", lwd=1.5)
    abline(h=c(.25,.5, .75), col="black", lty="dashed")

  }

}
par(mfrow=c(1, 1))

##############END CODE###########################
#################################################
```

The graph can be sorted 2 ways either by the untreated's decreasing relabund, or by the most abundant within each phylum. For each treatment, the resulting *C. difficile* CFU/g Feces is also labeled on the Y axis below the treatment name. File saved as "topdose2_tx1_sorted_10x20.pdf"

@@

### Figure 2 Correlation Calculations

@@

### Figure 2 Graph

>>>>

This figure was built using the R script called "abxD01.barcharts.xOTU.test.r" in the abxD01 folder.The input files are listed in Github. The graph can be sorted 2 ways either by the untreated's decreasing relabund, or by the most abundant within each phylum. For each treatment, the resulting *C. difficile* CFU/g Feces is also labeled on the Y axis below the treatment name. File saved as "topdose2_tx1_sorted_10x20.pdf" and "topdose2_tx2_sorted_10x20.pdf"."

@@

### Figure 2: Correlation Calculations

@@

### Figure 2: Stipchart
